---
title: "Draft EQI Models"
author: "Marie Rivers"
date: "11/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(kableExtra)
library(jtools)
library(ggstance)
library(huxtable)
library(tidymodels)
library(patchwork)
```

```{r}
eqi_pop <- read_csv(here("data", "eqi_pop.csv")) %>%
  filter(pop_change_pct < 100)
```

### Simple linear regression  

Linear regression models were used to evaluate the relationship between environmental quality and population change using the overall EQI value and each domain specific EQI to determine if a particular domain was a stronger predictor of population change.

```{r}
eqi_model <- lm(pop_change_pct ~ EQI, data = eqi_pop)
summary(eqi_model)
```

```{r}
air_eqi_model <- lm(pop_change_pct ~ air_EQI, data = eqi_pop)
water_eqi_model <- lm(pop_change_pct ~ water_EQI, data = eqi_pop)
land_eqi_model <- lm(pop_change_pct ~ land_EQI, data = eqi_pop)
built_eqi_model <- lm(pop_change_pct ~ built_EQI, data = eqi_pop)
sociodemographic_eqi_model <- lm(pop_change_pct ~ sociodemographic_EQI, data = eqi_pop)
```

```{r}
eqi_lm_plot <- ggplot(data = eqi_pop, aes(x = EQI, y = pop_change_pct)) + 
  geom_point(size = 0.5) + 
  geom_smooth(method = lm, formula = y~x, se = FALSE) +
  labs(y = "% pop change")

air_eqi_lm_plot <- ggplot(data = eqi_pop, aes(x = air_EQI, y = pop_change_pct)) + 
  geom_point(size = 0.5) + 
  geom_smooth(method = lm, formula = y~x, se = FALSE) +
  labs(y = "% pop change")

water_eqi_lm_plot <- ggplot(data = eqi_pop, aes(x = water_EQI, y = pop_change_pct)) + 
  geom_point(size = 0.5) + 
  geom_smooth(method = lm, formula = y~x, se = FALSE) +
  labs(y = "% pop change")

land_eqi_lm_plot <- ggplot(data = eqi_pop, aes(x = land_EQI, y = pop_change_pct)) + 
  geom_point(size = 0.5) + 
  geom_smooth(method = lm, formula = y~x, se = FALSE) +
  labs(y = "% pop change")

built_eqi_lm_plot <- ggplot(data = eqi_pop, aes(x = built_EQI, y = pop_change_pct)) + 
  geom_point(size = 0.5) + 
  geom_smooth(method = lm, formula = y~x, se = FALSE) +
  labs(y = "% pop change")

sociod_eqi_lm_plot <- ggplot(data = eqi_pop, aes(x = sociodemographic_EQI, y = pop_change_pct)) + 
  geom_point(size = 0.5) + 
  geom_smooth(method = lm, formula = y~x, se = FALSE) +
  labs(y = "% pop change")
```

```{r}
linear_model_plots <- (eqi_lm_plot | air_eqi_lm_plot | water_eqi_lm_plot) / (land_eqi_lm_plot | built_eqi_lm_plot | sociod_eqi_lm_plot )

linear_model_plots + plot_annotation(
  title = "Simple Linear Models of EQI Domains vs. % Population Change",
  subtitle = "US Counties, 2006-2010")
```

```{r}
plot_summs(eqi_model, air_eqi_model, water_eqi_model, land_eqi_model, built_eqi_model, sociodemographic_eqi_model, inner_ci_level = 0.9, model.names = c("EQI", "air EQI", "water EQI", "land EQI", "built EQI", "sociodem EQI"))
```

```{r}
export_summs(eqi_model, air_eqi_model, water_eqi_model, land_eqi_model, built_eqi_model, sociodemographic_eqi_model, 
             digits = 3, error_format = "[{conf.low}, {conf.high}]", 
             model.names = c("EQI", "air EQI", "water EQI", "land EQI", "built EQI", "sociodem EQI"))
```

### Linear regression with categorical variables

```{r}
eqi_cat_model <- lm(pop_change_pct ~ EQI + rucc_text, data = eqi_pop)
summary(eqi_cat_model)
```

```{r}
eqi_cat_plot_standalone <- ggplot(data = eqi_pop, aes(x = EQI, y = pop_change_pct, color = rucc_text)) +
  geom_point(size = 0.5) + 
  geom_line(data = augment(eqi_cat_model), aes(y = .fitted, color = rucc_text)) +
  theme(legend.position = "bottom") + 
  labs(title = "Categorical model of EQI vs. population change based on RUCC", y = "% population change", color = "rural-urban continual category:")
eqi_cat_plot_standalone
```

```{r}
summ(eqi_cat_model, confint = TRUE, digits = 3)
```

```{r}
plot_summs(eqi_cat_model, inner_ci_level = 0.9)
```

```{r}
air_eqi_cat_model <- lm(pop_change_pct ~ air_EQI + rucc_text, data = eqi_pop)
water_eqi_cat_model <- lm(pop_change_pct ~ water_EQI + rucc_text, data = eqi_pop)
land_eqi_cat_model <- lm(pop_change_pct ~ land_EQI + rucc_text, data = eqi_pop)
built_eqi_cat_model <- lm(pop_change_pct ~ built_EQI + rucc_text, data = eqi_pop)
sociodemographic_eqi_cat_model <- lm(pop_change_pct ~ sociodemographic_EQI + rucc_text, data = eqi_pop)
```

```{r}
eqi_cat_plot <- ggplot(data = eqi_pop, aes(x = EQI, y = pop_change_pct)) + 
  geom_point(aes(color = rucc_text), size = 0.5, show.legend = FALSE) + 
  geom_line(data = augment(eqi_cat_model), aes(y = .fitted, color = rucc_text), show.legend = FALSE) +
  labs(y = "% pop change")

air_eqi_cat_plot <- ggplot(data = eqi_pop, aes(x = air_EQI, y = pop_change_pct)) + 
  geom_point(aes(color = rucc_text), size = 0.5, show.legend = FALSE) + 
  geom_line(data = augment(air_eqi_cat_model), aes(y = .fitted, color = rucc_text), show.legend = FALSE) +
  labs(y = "% pop change")

water_eqi_cat_plot <- ggplot(data = eqi_pop, aes(x = water_EQI, y = pop_change_pct)) + 
  geom_point(aes(color = rucc_text), size = 0.5, show.legend = FALSE) + 
  geom_line(data = augment(water_eqi_cat_model), aes(y = .fitted, color = rucc_text), show.legend = FALSE) +
  labs(y = "% pop change")

land_eqi_cat_plot <- ggplot(data = eqi_pop, aes(x = land_EQI, y = pop_change_pct)) + 
  geom_point(aes(color = rucc_text), size = 0.5, show.legend = FALSE) + 
  geom_line(data = augment(land_eqi_cat_model), aes(y = .fitted, color = rucc_text), show.legend = FALSE) +
  labs(y = "% pop change")

built_eqi_cat_plot <- ggplot(data = eqi_pop, aes(x = built_EQI, y = pop_change_pct)) + 
  geom_point(aes(color = rucc_text), size = 0.5) + 
  geom_line(data = augment(built_eqi_cat_model), aes(y = .fitted, color = rucc_text), show.legend = FALSE) +
  labs(y = "% pop change", color = "RUCC:") +
  theme(legend.position = "bottom")

sociod_eqi_cat_plot <- ggplot(data = eqi_pop, aes(x = sociodemographic_EQI, y = pop_change_pct)) + 
  geom_point(aes(color = rucc_text), size = 0.5, show.legend = FALSE) + 
  geom_line(data = augment(sociodemographic_eqi_cat_model), aes(y = .fitted, color = rucc_text), show.legend = FALSE) +
  labs(y = "% pop change", x = "sociodem_EQI")
```

```{r}
cat_model_plots <- (eqi_cat_plot | air_eqi_cat_plot | water_eqi_cat_plot) / (land_eqi_cat_plot | built_eqi_cat_plot| sociod_eqi_cat_plot )

cat_model_plots + plot_annotation(
  title = "Categorical Linear Models of EQI Domains vs. % Population Change",
  subtitle = "US Counties, 2006-2010")
```

```{r}
plot_summs(eqi_cat_model, air_eqi_cat_model, water_eqi_cat_model, land_eqi_cat_model, built_eqi_cat_model, sociodemographic_eqi_cat_model, inner_ci_level = 0.9, model.names = c("EQI", "air EQI", "water EQI", "land EQI", "built EQI", "sociodem EQI"))
```

```{r}
export_summs(eqi_cat_model, air_eqi_cat_model, water_eqi_cat_model, land_eqi_cat_model, built_eqi_cat_model, sociodemographic_eqi_cat_model, 
             digits = 3, error_format = "[{conf.low}, {conf.high}]", 
             model.names = c("EQI", "air EQI", "water EQI", "land EQI", "built EQI", "sociodem EQI"))
```

### Interaction model  

Linear regression was used to explore the relationship between EQI and population change. Looking at the data, the relationship between EQI and population change may be influenced by the rural-urban continuum of each county. The interaction model used to describe this relationship is:

$$\text{percent population change}_i=\beta_{0}+\beta_{1} \cdot EQI_i + \beta_{2} \cdot RUCC_i + \beta_{3} \cdot EQI_i \cdot RUCC_i + \varepsilon_i$$

```{r}
eqi_int_model <- lm(pop_change_pct ~ EQI + rucc_text + EQI:rucc_text, data = eqi_pop)
summary(eqi_int_model)
```

```{r}
eqi_int_model_plot_standalone <- ggplot(data = eqi_pop, aes(x = EQI, y = pop_change_pct, color = rucc_text)) +
  geom_point(size = 0.5) + 
  geom_smooth(method = "lm", se = FALSE) +
  theme(legend.position = "bottom") + 
  labs(title = "Interaction model of EQI vs. population change based on RUCC", y = "% population change", color = "rural-urban continual category:")
eqi_int_model_plot_standalone
```

```{r}
summ(eqi_int_model, confint = TRUE, digits = 3)
```

```{r}
plot_summs(eqi_int_model, inner_ci_level = 0.9)
```

Additionally, interaction models were created for each domain specific EQI to determine if a particular domain was more strongly correlated with population change.

```{r}
air_eqi_int_model <- lm(pop_change_pct ~ air_EQI + rucc_text + air_EQI:rucc_text, data = eqi_pop)
water_eqi_int_model <- lm(pop_change_pct ~ water_EQI + rucc_text + water_EQI:rucc_text, data = eqi_pop)
land_eqi_int_model <- lm(pop_change_pct ~ land_EQI + rucc_text + land_EQI:rucc_text, data = eqi_pop)
built_eqi_int_model <- lm(pop_change_pct ~ built_EQI + rucc_text + built_EQI:rucc_text, data = eqi_pop)
sociodemographic_eqi_int_model <- lm(pop_change_pct ~ sociodemographic_EQI + rucc_text + sociodemographic_EQI:rucc_text, data = eqi_pop)
```

```{r}
plot_summs(eqi_int_model, air_eqi_int_model, water_eqi_int_model, land_eqi_int_model, built_eqi_int_model, sociodemographic_eqi_int_model, inner_ci_level = 0.9, model.names = c("EQI", "air EQI", "water EQI", "land EQI", "built EQI", "sociodem EQI"))
```

```{r}
export_summs(eqi_int_model, air_eqi_int_model, water_eqi_int_model, land_eqi_int_model, built_eqi_int_model, sociodemographic_eqi_int_model, 
             digits = 3, error_format = "[{conf.low}, {conf.high}]", 
             model.names = c("EQI", "air EQI", "water EQI", "land EQI", "built EQI", "sociodem EQI"))
```
```{r}
# eqi_int_model_plot_standalone <- ggplot(data = eqi_pop, aes(x = EQI, y = pop_change_pct, color = rucc_text)) +
#   geom_point(size = 0.5) + 
#   geom_smooth(method = "lm", se = FALSE) +
#   theme(legend.position = "bottom") + 
#   labs(title = "Interaction model of EQI vs. population change based on RUCC", y = "% population change", color = "rural-urban continual category:")

eqi_int_plot <- ggplot(data = eqi_pop, aes(x = EQI, y = pop_change_pct, color = rucc_text)) + 
  geom_point(size = 0.5, show.legend = FALSE) + 
  geom_smooth(method = "lm", se = FALSE, show.legend = FALSE) +
  labs(y = "% pop change")

air_eqi_int_plot <- ggplot(data = eqi_pop, aes(x = air_EQI, y = pop_change_pct, color = rucc_text)) + 
  geom_point(size = 0.5, show.legend = FALSE) +  
  geom_smooth(method = "lm", se = FALSE, show.legend = FALSE) +
  labs(y = "% pop change")

water_eqi_int_plot <- ggplot(data = eqi_pop, aes(x = water_EQI, y = pop_change_pct, color = rucc_text)) + 
  geom_point(size = 0.5, show.legend = FALSE) + 
  geom_smooth(method = "lm", se = FALSE, show.legend = FALSE) +
  labs(y = "% pop change")

land_eqi_int_plot <- ggplot(data = eqi_pop, aes(x = land_EQI, y = pop_change_pct, color = rucc_text)) + 
  geom_point(size = 0.5, show.legend = FALSE) + 
  geom_smooth(method = "lm", se = FALSE, show.legend = FALSE) +
  labs(y = "% pop change")

built_eqi_int_plot <- ggplot(data = eqi_pop, aes(x = built_EQI, y = pop_change_pct, color = rucc_text)) + 
  geom_point(size = 0.5) + 
  geom_smooth(method = "lm", se = FALSE, show.legend = FALSE) +
  labs(y = "% pop change", color = "RUCC:") +
  theme(legend.position = "bottom")

sociod_eqi_int_plot <- ggplot(data = eqi_pop, aes(x = sociodemographic_EQI, y = pop_change_pct, color = rucc_text)) + 
  geom_point(size = 0.5, show.legend = FALSE) +  
  geom_smooth(method = "lm", se = FALSE, show.legend = FALSE) +
  labs(y = "% pop change", x = "sociodem_EQI")
```

```{r}
int_model_plots <- (eqi_int_plot | air_eqi_int_plot | water_eqi_int_plot) / (land_eqi_int_plot | built_eqi_int_plot| sociod_eqi_int_plot )

int_model_plots + plot_annotation(
  title = "Interaction Models of EQI Domains vs. % Population Change",
  subtitle = "US Counties, 2006-2010")
```
